$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-09-08 paper/paper.tex
Bring lemmas forward to the body

% Do not use italics in definitions and theorems
\theoremstyle{definition}
\newtheorem{nidefinition}{Definition}
\newtheorem{nitheorem}{Theorem}
\newtheorem{nilemma}{Lemma}
\renewcommand{\sectionautorefname}{\S}%
\renewcommand{\subsectionautorefname}{\S}%

@@

In rule $\constr{comm}$, parallel processes reduce when they communicate over a common channel with index ${i}$.
As a result of that communication, the continuation of the input process $P$ has all the references to its most immediate variable substituted with references to $\suc j$, the variable sent by the output process $\send{i}{j}{Q}$.
After this substitution, $\subst{P}{\suc j}{\constr{0}}$ is \emph{lowered} --- all variable references are decreased by one (we apply \autoref{lm:substitution-unused} in \autoref{app:lemmas} to obtain a proof $\Unused_{\constr{0}} \; (\subst{P}{\suc j}{\constr{0}})$).
Reduction is closed under parallel composition (rule $\constr{par}$), restriction (rule $\constr{res}$) and structural congruence (rule $\constr{struct}$) 
--- notably, not under input nor output, as doing so would not preserve the sequencing of actions \cite{Sangio01}.

@@

\subsection{Towards Subject Reduction}
\paragraph*{Subject Congruence}
states that applying structural congruence rules to a well typed process preserves its well-typedness.

@@

\paragraph*{Subject Reduction}
states that if $P$ is well typed and it reduces to $Q$, then $Q$ is well typed.
In the \picalc{} we distinguish between a reduction $P \reduce{\constr{internal}} Q$ on a channel internal to $P$, and a reduction $P \reduce{\constr{external} \; i} Q$ on a channel $i$ external to $P$ (refer to \autoref{operational-semantics}).

>>>>

% Do not use italics in definitions and theorems
\theoremstyle{definition}
\newtheorem{nidefinition}{Definition}
\newtheorem{nitheorem}{Theorem}
\newtheorem{nilemma}{Lemma}
\renewcommand{\nidefinitionautorefname}{Definition}%
\renewcommand{\nitheoremautorefname}{Theorem}%
\newcommand{\nilemmaautorefname}{Lemma}%
\renewcommand{\sectionautorefname}{\S}%
\renewcommand{\subsectionautorefname}{\S}%

@@

In rule $\constr{comm}$, parallel processes reduce when they communicate over a common channel with index ${i}$.
As a result of that communication, the continuation of the input process $P$ has all the references to its most immediate variable substituted with references to $\suc j$, the variable sent by the output process $\send{i}{j}{Q}$.
After this substitution, $\subst{P}{\suc j}{\constr{0}}$ is \emph{lowered} --- all variable references are decreased by one (we apply \autoref{lm:substitution-unused} to obtain a proof $\Unused_{\constr{0}} \; (\subst{P}{\suc j}{\constr{0}})$).
Reduction is closed under parallel composition (rule $\constr{par}$), restriction (rule $\constr{res}$) and structural congruence (rule $\constr{struct}$) 
--- notably, not under input nor output, as doing so would not preserve the sequencing of actions \cite{Sangio01}.

\begin{nilemma}
  \label{lm:substitution-unused}
  For every variables $i$ and $j$, if $i \type{\not\equiv} j$ then $\Unused_i \allowbreak (\subst{P}{j}{i})$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\subsection{Towards Subject Reduction}
\paragraph*{Subject Congruence}
states that applying structural congruence rules to a well typed process preserves its well-typedness.
We must first introduce lemmas that establish that some syntactic manipulations can be inverted and how unused variables relate to the preservation of leftovers.

\begin{nilemma}
  \label{lm:lower-lift}
  The function $\func{lower}_i \; P \; uP$ has an inverse $\func{lift}_i \; P$ that increments every $\textsc{Var}$ greater than or equal to $i$, such that $\func{lift}_i \; (\func{lower}_i \; P \; uP) \equiv P$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{nilemma}
  \label{lm:exchange-exchange}
  The function $\func{exchange}_i \; P$ is its own inverse: $\func{exchange}_i \; (\func{exchange}_i \; P) \equiv P$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{nilemma}
  \label{lm:types-unused}
  For all well-typed processes $\types{\gamma}{\Gamma}{P}{\Xi}$, if the variable $i$ is unused within $P$, then $\Gamma$ at $i$ is equivalent to $\Xi$ at $i$.
\end{nilemma}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

We are now in a position to prove subject congruence.

@@

\paragraph*{Subject Reduction}
states that if $P$ is well typed and it reduces to $Q$, then $Q$ is well typed.
In the \picalc{} we distinguish between a reduction $P \reduce{\constr{internal}} Q$ on a channel internal to $P$, and a reduction $P \reduce{\constr{external} \; i} Q$ on a channel $i$ external to $P$ (refer to \autoref{operational-semantics}).
We first introduce a helper lemma:

\begin{nilemma}
  \label{lm:comm-capable}
  Every input usage context $\Gamma$ of a well-typed process $\types{\gamma}{\Gamma}{P}{\Delta}$ that reduces by communicating on a channel external (that is, $P \reduce{\constr{external} \; i} Q$ for some $Q$) has a multiplicity of at least $\lio$ at index $i$.
\end{nilemma}

\begin{proof}
  By induction on the reduction derivation $P \reduce{\constr{external \; i}}Q$.
\end{proof}
